\input{wg21common}

% Footnotes at bottom of page:
 \usepackage[bottom]{footmisc} 

% Table going across a page: 
 \usepackage{longtable}

 % Start sections at 0
% \setcounter{section}{-1}

% color boxes
\usepackage{tikz,lipsum,lmodern}
\usepackage[most]{tcolorbox}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%TABLE OF CONTENTS SETTINGS

\usepackage{titlesec}
\usepackage{tocloft}

% Custom ToC layout because the default sucks
\cftsetindents{section}{0in}{0.24in}
\cftsetindents{subsection}{0.24in}{0.34in}
\cftsetindents{subsubsection}{0.58in}{0.44in}

% Needed later to reduce the ToC depth mid document
\newcommand{\changelocaltocdepth}[1]{%
  \addtocontents{toc}{\protect\setcounter{tocdepth}{#1}}%
  \setcounter{tocdepth}{#1}%
}

\setcounter{tocdepth}{3}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%POLLS

\definecolor{pollFrame}{rgb}{0,.718,0}
\definecolor{pollBG}{rgb}{.5,1,.5}

\newtcolorbox{wgpoll}[1]{colframe=pollFrame,colback=pollBG!20!white,title={#1}}

\newcommand{\wgpollresult}[5]{%

  \vspace{2mm}
  \begin{tabular}{c | c | c | c | c} %
  SF  & F  & N  & A  & SA \\ %
  \hline %
  #1 & #2 & #3 & #4 & #5 \\ %

  \end{tabular}
  \vspace{2mm}  \\ %
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}
\title{Require a non-throwing default contract-violation handler}
\author{
John Lakos \small(\href{mailto:jlakos@bloomberg.net}{jlakos@bloomberg.net}) 
}
\date{}
\maketitle

\begin{tabular}{ll}
Document \#: & P3577R0 \\
Date: &2025-01-13 \\
Project: & Programming Language C++ \\
Audience: & SG21, EWG
\end{tabular}

\begin{abstract}
This paper proposes a one-line addition to require that the system-supplied default violation handler not throw.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\tableofcontents*
%\pagebreak

%\section*{Revision history}

%Revision 0 (2024-04-16)
%\begin{itemize}
%\item Original version
%\end{itemize}

%\pagebreak

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\section{Introduction}

Some people are concerned that a new codepath might result if an \emph{implicit} precondition --- a precondition attached by the implementation to a core language expression, as proposed in \cite{P3081R1}, \cite{P3100R1}, and \cite{P3329R0} --- is violated and the contract-violation handler exits via an exception.

This paper addresses that problem by requiring that the \emph{default} contract-violation handler does not exit via an exception. Thus, the new code path will not happen without an explicit replacement of the contract-violation handler by the user, which means it cannot happen inadvertently by accident or as the result of a port to a conforming implementation.

Note that this restriction can always be relaxed in a later iteration of C++ as a non-breaking change, but it cannot be so added.

\section{Proposed wording}

The proposed wording is relative to \cite{P2900R13}.

Modify [basic.contract.handler], paragraph 1 as follows:

\begin{adjustwidth}{0.5cm}{0.5cm}
The \emph{contract-violation handler} of a program is a function named \mbox{\tcode{::handle_contract_violation}}. The contract-violation handler shall take a single argument of type ``lvalue reference to \tcode{const} \mbox{\tcode{std::contracts::contract_violation}}” and shall return \tcode{void}. The contract-violation handler may have a non-throwing exception specification. The implementation shall provide a definition of the contract-violation handler, called the \emph{default contract-violation handler.}\added{ The default contract-violation handler shall not exit via an exception.}
\begin{note}
No declaration for the default contract-violation handler is provided by any standard library header.
\end{note}
\end{adjustwidth}

\section{Summary}

By adding this one line of wording, we eliminate a stated concern, and leave open all options moving forward. It is anticipated that this proposal will increase consensus, as it has no obvious or likely negative consequences.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Acknowledgements}

Thanks to Timur Doumler for providing the wording.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Remove ToC entry for bibliography
\renewcommand{\addcontentsline}[3]{}% Make \addcontentsline a no-op to disable auto ToC entry

%\renewcommand{\bibname}{References}  % custom name for bibliography
\bibliographystyle{abstract}
\bibliography{ref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\end{document}
